---
title: "my title"
author: "Dummy Surname (dummy@mail.com)"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"      
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 1

**Build a Random Forest model to classify COVID-19 severity**

Using the training cohort provided in Supplementary Table 3, build a Random Forest classification model similar to the one in the publication to distinguish between non-Severe and Severe COVID-19 patients.

Report the following performance metrics from cross-validation:

-   Confusion matrix\
-   Accuracy\
-   ROC curve and AUC

How well does the Random Forest model separate non-Severe from Severe patients based on proteomic profiles?

```{r}

# --- STEP 1: LIBRARIES ---
library(readxl)
library(tidyverse)
library(janitor)
library(httr)
library(tidymodels)
library(ranger)
library(vip)

# --- STEP 2: DOWNLOAD DATA ---

# 1. Download Table 3 (Training Proteomics)
mmc3_url <- "https://ars.els-cdn.com/content/image/1-s2.0-S0092867420306279-mmc3.xlsx"
mmc3_file <- tempfile(fileext = ".xlsx")
GET(mmc3_url, write_disk(mmc3_file, overwrite = TRUE))

df_mmc3_raw <- read_excel(mmc3_file, sheet = 2, skip = 1, na = c("", "NA", "/"))

# 2. Download Metadata (Labels)
meta_url <- "https://ars.els-cdn.com/content/image/1-s2.0-S0092867420306279-mmc1.xlsx"
meta_file <- tempfile(fileext = ".xlsx")
GET(meta_url, write_disk(meta_file, overwrite = TRUE))

df_metadata_raw <- read_excel(meta_file, sheet = 2, na = c("", "NA", "/"))

# --- STEP 3: CLEANING & FIXING THE IDs ---

# A. Clean Table 3
df_mmc3 <- df_mmc3_raw %>%
  clean_names() %>%
  rename(protein = 1) %>% 
  select(-2) # Gene Symbol entfernen

# Transponieren
df_mmc3_transposed <- df_mmc3 %>%
  column_to_rownames(var = "protein") %>%
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Patient_ID") %>%
  # --- FIX: IDs GROSS SCHREIBEN ---
  mutate(Patient_ID = toupper(Patient_ID)) 

# B. Clean Metadata
df_labels <- df_metadata_raw %>%
  clean_names() %>%
  mutate(
    group = case_when(
      group_d == 0 ~ "Non-COVID-19",
      group_d == 1 ~ "Healthy",
      group_d == 2 ~ "Non-severe",
      group_d == 3 ~ "Severe",
      TRUE ~ NA_character_
    ),
    group = factor(group, levels = c("Healthy", "Non-COVID-19", "Non-severe", "Severe")),
    # --- FIX: IDs GROSS SCHREIBEN & LEERZEICHEN WEG ---
    patient_id_a = toupper(str_trim(patient_id_a))
  ) %>% 
  select(patient_id_a, group)

# --- STEP 4: MERGE (Training Data vorbereiten) ---

# Jetzt sollte der Join klappen, da beide IDs "XG1" heißen
ml_data <- df_mmc3_transposed %>%
  inner_join(df_labels, by = c("Patient_ID" = "patient_id_a")) %>%
  filter(group %in% c("Severe", "Non-severe")) %>%
  mutate(group = droplevels(group)) %>%
  select(-Patient_ID)

print("--- Data Check (MUSS > 0 sein!) ---")
print(table(ml_data$group))
# Erwartet: ca. 26 Severe, 39 Non-severe

# --- STEP 5: MACHINE LEARNING (Exercise 1) ---

set.seed(123)

# 1. Split (Wir teilen Table 3 nochmal intern zum Validieren)
data_split <- initial_split(ml_data, prop = 0.70, strata = group)
train_data <- training(data_split)
test_data  <- testing(data_split)

# 2. Recipe
rf_recipe <- recipe(group ~ ., data = train_data) %>%
  step_filter_missing(all_predictors(), threshold = 0.4) %>% 
  step_impute_knn(all_numeric_predictors(), neighbors = 5) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

# 3. Model
rf_spec <- rand_forest(trees = 500) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

# 4. Workflow
rf_workflow <- workflow() %>%
  add_recipe(rf_recipe) %>%
  add_model(rf_spec)

# 5. Training & Evaluation
print("Starting Random Forest Training...")
final_res <- last_fit(rf_workflow, data_split)

# --- RESULTS ---

print("--- Accuracy & AUC ---")
print(collect_metrics(final_res))

print("--- Confusion Matrix ---")
preds <- collect_predictions(final_res)
print(conf_mat(preds, truth = group, estimate = .pred_class))

print("--- ROC Curve ---")
roc_curve(preds, truth = group, .pred_Severe) %>% 
  autoplot() + 
  ggtitle("ROC Curve: Training Cohort (Internal Validation)") +
  theme_minimal() 
```

# Exercise 2

**Identify and interpret the most important features**

Using the trained Random Forest model from Exercise 1, identify the 25 most important protein features contributing to the classification. Disccus their biological relevance.

Compare your selected features to the proteins reported by the authors in Supplementary Table 5.

```{r}

```

# Exercise 3

**External prediction on an independent test cohort**

Using the final Random Forest model trained on the full training cohort, predict disease severity for patients in the independent test cohort provided in Supplementary Table 4.

Your task is to:

-   Apply the trained model to the test cohort.  
-   Obtain predicted probabilities of Severe disease for each patient.  
-   Assign predicted class labels based on these probabilities.  
-   Compare predictions to the true clinical labels.  


How does the model perform on the independent test cohort?

```{r}

```

# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
